Calculating life expectancy and exploring missingness

### 2016-11-08 Code/1_Explore_Data_v2.Rmd
computed LE and decomp for 1 and 5

```{r calculate_life_expectancy}
list.life.tables <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
#list.life.tables1 <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
list.index <- 1

@@

```{r add_le_to_dataset}
le <- rep(NA, length(non.zero.stratums))
le5 <- rep(NA, length(non.zero.stratums))
le9 <- rep(NA, length(non.zero.stratums))

@@

rm(les)
```

@@

Black <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth, life.expectancy.birth5, life.expectancy.birth9) %>%
                  rename(Pop_black = Pop_across_age,  
                         le.birth.black = life.expectancy.birth, 
                         le.birth.black5 = life.expectancy.birth5, 
                         le.birth.black9 = life.expectancy.birth9)

White <- dat2 %>% filter(Race2 == "White" & Age3 == 0) %>% 
                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth, life.expectancy.birth5, life.expectancy.birth9) %>%
                  rename(Pop_white = Pop_across_age,  
                         le.birth.white = life.expectancy.birth,
                         le.birth.white5 = life.expectancy.birth5,
                         le.birth.white9 = life.expectancy.birth9)

@@

    list.age.decomp.tables9[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables9[[black.index]]), 
                                                 name.lt1 = "Black", 
                                                 life.table2 = as.data.frame(list.life.tables9[[white.index]]), 
                                                 name.lt2 = "White", age.groups = "Age3")
  } else {
    list.age.decomp.tables[[i]] <- list.age.decomp.tables5[[i]] <- list.age.decomp.tables9[[i]] <-"Couldn't compare life tables"
  }
}
```

@@

#takes about 20 mins
list.cod.decomp.tables <- list.cod.decomp.tables5 <- list.cod.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])

@@

    list.cod.decomp.tables[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
                                                         life.table2 = as.data.frame(list.life.tables[[white.index]]), 
                                                         decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
                                                         cod.table = ct.both %>% select(Age2, COD2, prop.white5, prop.black5), 
                                                         age.colname.cod.table = "Age2", 
                                                         COD.colname.cod.table = "COD2", 
                                                         prop1.colname.cod.table = "prop.black", 
                                                         prop2.colname.cod.table = "prop.white")
  } else {
    list.cod.decomp.tables[[i]] <- "Couldn't compare life tables"
  }  
  #print(i)
}
```

```{r  update_decomp_for_plotting}
#update the decomposition table for plotting purposes
for(i in 1:dim(paired.ids)[1]) {
    if(list.cod.decomp.tables[[i]][1] != "Couldn't compare life tables") { 
        list.cod.decomp.tables[[i]] <- 
                              make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables[[i]], 
                              age.groups = "Ages", 
                              cause.of.death = "Cause.of.death")  
    } 
  #print(i)
}  
```

```{r pretty_cod}
ggplotly(ggplot(list.cod.decomp.tables[[1]], aes(y = Ages, x = start)) + 
           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
           xlab("Contribution to life expectancy gap") +
           ylab("Age group\n") + theme_minimal() +
           geom_vline(xintercept = 0))
```

@@

```{r save_specified_files}
save(dat.clean, dat2, BlackWhite, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
     file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
```

>>>>

```{r calculate_life_expectancy}
#takes 10 mins for all three

list.life.tables <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
#list.life.tables1 <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
list.index <- 1

@@

```{r add_le_to_dataset}
le <- le5 <- le9 <- rep(NA, length(non.zero.stratums))

@@

rm(les)

ggplot(dat2 %>% filter(Age2 == "<1 year"), aes(x = life.expectancy.birth)) + 
  geom_density(aes(fill="Impute 1"), alpha = 0.3) + 
  geom_density(aes(x = life.expectancy.birth.5, fill = "Impute 5"), alpha = 0.3) +
  geom_density(aes(x = life.expectancy.birth.9, fill="Impute 9"), alpha = 0.3) +
  facet_grid(Sex2 ~ Race2)
```

@@

Black <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
                  select(State2, Year3, Sex2, Pop_across_age, 
                         life.expectancy.birth, life.expectancy.birth.5, 
                         life.expectancy.birth.9) %>%
                  rename(Pop_black = Pop_across_age,  
                         le.birth.black = life.expectancy.birth, 
                         le.birth.black5 = life.expectancy.birth.5, 
                         le.birth.black9 = life.expectancy.birth.9)

White <- dat2 %>% filter(Race2 == "White" & Age3 == 0) %>% 
                  select(State2, Year3, Sex2, Pop_across_age, 
                         life.expectancy.birth, life.expectancy.birth.5, 
                         life.expectancy.birth.9) %>%
                  rename(Pop_white = Pop_across_age,  
                         le.birth.white = life.expectancy.birth,
                         le.birth.white5 = life.expectancy.birth.5,
                         le.birth.white9 = life.expectancy.birth.9)

@@

    list.age.decomp.tables9[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables9[[black.index]]), 
                                                 name.lt1 = "Black", 
                                                 life.table2 = as.data.frame(list.life.tables9[[white.index]]), 
                                                 name.lt2 = "White", age.groups = "Age3")
  } else {
    list.age.decomp.tables[[i]] <- list.age.decomp.tables5[[i]] <- list.age.decomp.tables9[[i]] <-"Couldn't compare life tables"
  }
}

@@

#takes 1.5 hours for all three
list.cod.decomp.tables <- list.cod.decomp.tables5 <- list.cod.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])

system.time(
for(i in 1:dim(paired.ids)[1]) {
  white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
  black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])

@@

    if(any(is.na(ct.both$prop.black1)) == F & any(is.na(ct.both$prop.white1)) == F) { #ensures that there are no ages with 0 deaths - for these COD prop == 0 and can't compute

        list.cod.decomp.tables[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white1, prop.black1), 
                                                              age.colname.cod.table = "Age2", 
                                                              COD.colname.cod.table = "COD2", 
                                                              prop1.colname.cod.table = "prop.black1", 
                                                              prop2.colname.cod.table = "prop.white1")

        list.cod.decomp.tables5[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white5, prop.black5), 
                                                              age.colname.cod.table = "Age2", 
                                                              COD.colname.cod.table = "COD2", 
                                                              prop1.colname.cod.table = "prop.black5", 
                                                              prop2.colname.cod.table = "prop.white5")

        list.cod.decomp.tables9[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white9, prop.black9), 
                                                              age.colname.cod.table = "Age2", 
                                                              COD.colname.cod.table = "COD2", 
                                                              prop1.colname.cod.table = "prop.black9", 
                                                              prop2.colname.cod.table = "prop.white9")

    } else { list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Encountered age group with no deaths"

    }
  } else {

    list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Couldn't compare life tables"
  }

  print(i)
}
)
```

```{r  update_decomp_for_plotting}
#update the decomposition table for plotting purposes
for(i in 1:dim(paired.ids)[1]) {
    if(list.cod.decomp.tables[[i]][1] != "Couldn't compare life tables" & 
       list.cod.decomp.tables[[i]][1] != "Encountered age group with no deaths") {

       list.cod.decomp.tables[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables[[i]],
                                                            age.groups = "Ages", 
                                                            cause.of.death = "Cause.of.death")

       list.cod.decomp.tables5[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables5[[i]],
                                                            age.groups = "Ages", 
                                                            cause.of.death = "Cause.of.death")

       list.cod.decomp.tables9[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables9[[i]],
                                                            age.groups = "Ages", 
                                                            cause.of.death = "Cause.of.death") 
    } 
  print(i)
}  
```

```{r pretty_cod}
ggplotly(ggplot(list.cod.decomp.tables[[1]], aes(y = Ages, x = start)) + 
           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
           xlab("Contribution to life expectancy gap") +
           ylab("Age group\n") + theme_minimal() +
           geom_vline(xintercept = 0))

ggplotly(ggplot(list.cod.decomp.tables9[[1]], aes(y = Ages, x = start)) + 
           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
           xlab("Contribution to life expectancy gap") +
           ylab("Age group\n") + theme_minimal() +
           geom_vline(xintercept = 0))
```

@@

```{r save_specified_files}
save(dat.clean, dat2, BlackWhite, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
     list.age.decomp.tables5, list.cod.decomp.tables5, list.age.decomp.tables9, list.cod.decomp.tables9,
     file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
```
